### 2023-07-29 2023-NHA-TRANG-Data-Analytics-for-Managers/Files-Trinh-Bay/Slide/04-tidydata.Rmd
update

---
title     : "Các Nguyên Lý Cơ Bản Trong Phân Tích Dữ Liệu"
subtitle  : "Làm Sạch và Hợp Thức Hóa Dữ Liệu, Dữ Liệu Gọn Gàng"
author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
date      : "4-tháng 8-2023"
#date      : "`r format(Sys.time(), '%d %B, %Y')`"
output    : 
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Bản Quyền của Cty TNHH ASQ Việt Nam - Chuẩn Bị cho Khóa Học tại Đại Học UEH"
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

## Data Cleansing and Validation (1)

> Data cleansing or data cleaning is the process of detecting and correcting (or removing) corrupt or inaccurate records from a record set, table, or database and refers to identifying incomplete, incorrect, inaccurate or irrelevant parts of the data and then replacing, modifying, or deleting the dirty or coarse data. Data cleansing may be performed interactively with data wrangling tools, or as batch processing through scripting
>
> `r tufte::quote_footer('--- https://en.wikipedia.org/wiki/Data_cleansing')`

<span style="color: red;"> Data scientists spend **60% of their time on cleaning** and organizing data. Collecting data sets comes second at 19% of their time, meaning data scientists spend around **80% of their time on preparing and managing data for analysis**.</span>

> Data Validation is an activity verifying whether or not a combination of values is a member of a set of acceptable combinations
>
> `r tufte::quote_footer('--- [Zio et al. 2016](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)')`

## Data Cleansing and Validation (2)

@@

Source: [Methodology for Data Valication 1.0](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)

@@

- Validation Level 0: consistency with the expected IT structural requirements

- Validation Level 1: consistency within the data set
- Validation Level 2:  consistency with other data sets within the same domain and within the same data source
- Validation Level 3:  consistency within the same domain between different data sources
- Validation Level 4:  consistency between separate domains in the same data provider 
- Validation Level 5: consistency with data of other data providers.

@@

## Tidy Data - Structure

@@

**Untidy**

Typical representation

@@

Another typical representation

@@

- What are the <span style="color: red;">**variable**</span>, <span style="color: red;">**observation**</span>, <span style="color: red;">**value**</span>?

@@

<div class='right' style='float:right;width:48%'>
> Happy families are all alike; every unhappy family is unhappy in its own way
>
> `r tufte::quote_footer('--- Leo Tolstoy')`

Like families, tidy datasets are all alike but every messy dataset is messy in its own way. Tidy datasets provide a standardized way to link the structure of a dataset (its physical layout) with its semantics (its meaning).

@@

## Tidy Data - Definition

@@

A dataset is a collection of <span style="color: red;">**values**</span>, usually either numbers (if quantitative) or strings (if qualitative). Values are organised in two ways. Every value belongs to a <span style="color: red;">**variable**</span> and an <span style="color: red;">**observation**</span>. A variable contains all values that measure the same underlying attribute (like height, temperature, duration) across units. An observation contains all values measured on the same unit (like a person, or a day, or a race) across attributes.

@@

source: [Openscapes](https://www.openscapes.org/blog/2020/10/12/tidy-data/)
</div>

@@

**Tidy data is a standard way of mapping the meaning of a dataset to its structure. A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types.** In tidy data:

  - Each variable forms a column.
  - Each observation forms a row.
  - Each type of observational unit forms a table.

@@

## Tidy Data - Tydy Our Room with Marie Kondo's Way (KONMARI)

@@

Folding a perfect underwear drawer

@@

Organize the world: Design Your Life to Spark Joy

@@

## Tidying Messy Data (1)

<div class='left' style='float:left;width:48%'>
- Column headers are values, not variable names.

- Multiple variables are stored in one column.

- Variables are stored in both rows and columns.

- Multiple types of observational units are stored in the same table.

- A single observational unit is stored in multiple tables
</div>

@@

```{r echo=FALSE, message=FALSE,warning=FALSE}
library(readxl)
library(dplyr)
# library(ggplot2)
data=read.csv("../data/tidy-data/data/pew-raw.csv", sep=",", header=TRUE)
data <-data%>%select(c(2:8))
#colnames(data) <- c('religion','<$10k','$10-20k','$20-30k','$30-40k','$40-50k','$50-75k')
library(DT)
p<-datatable(data,rownames = FALSE)
p
#library(knitr)
#library(kableExtra)
#data %>% 
#  kbl() %>%
#  kable_classic(full_width = F, html_font = "Cambria")
#  kable_paper(bootstrap_options = "striped", full_width = F,font_size = 22)
#kable(data, booktabs = TRUE)
```

<div class='right' style='float:left;width:15%'>

```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
knitr::include_graphics(url)
```

- Variables

  - religion
  - income
  - frequency

<div class='right' style='float:left;width:30%'>

Tidy Data

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../code/pew-tidying.R")
library(DT)
p<-datatable(molten.df,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left', targets = c(1,2)))
            ,autoWidth = TRUE)
             )
p
```

## Column headers are values, not variable names (2)

Billboard Ranking

```{r echo=FALSE, message=FALSE,warning=FALSE}
df=read.csv("../data/tidy-data/data/billboard.csv", sep=",", header=TRUE)
p<-datatable(df,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Column headers are values, not variable names (3)

Billboard Ranking

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
p<-datatable(clean,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left', targets = c(1,2,3,4,5,6)))
            ,autoWidth = TRUE)
             )
p
```

## Multiple variables stored in one column (1)

WHO data on turboculis (TB)

```{r echo=FALSE, message=FALSE,warning=FALSE}
df=read.csv("../data/tidy-data/data/tb.csv", sep=",", header=TRUE)
p<-datatable(df,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Multiple variables stored in one column (2)

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/tb.r")
library(DT)
p<-datatable(raw,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Multiple variables stored in one column (3)

<div class='right' style='float:left;width:45%'>

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/tb.r")
library(DT)
p<-datatable(clean1,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```

<div class='right' style='float:left;width:10%'>
```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
knitr::include_graphics(url)
```
</div>

<div class='right' style='float:left;width:45%'>
```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/tb.r")
library(DT)
p<-datatable(clean,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```
</div>

## Variables are stored in both rows and columns (1)

Global Historical Climatology Network for one weather station (MX17004)

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/weather.r")
library(DT)
p<-datatable(raw,rownames = FALSE,extensions = c('FixedColumns'),
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE,scrollX = TRUE,fixedColumns = TRUE)
             )
p
```

## Variables are stored in both rows and columns (2)

<div class='right' style='float:left;width:45%'>

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/weather.r")
library(DT)
p<-datatable(clean1,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```

<div class='right' style='float:left;width:10%'>
```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://hotemoji.com/images/dl/i/black-rightwards-arrow-emoji-by-twitter.png"
knitr::include_graphics(url)
```
</div>

<div class='right' style='float:left;width:45%'>
```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/weather.r")
library(DT)
p<-datatable(clean2,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```
</div>

## Multiple types in one table (1)

Billboard Ranking

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
p<-datatable(clean,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left', targets = c(1,2,3,4,5,6)))
            ,autoWidth = TRUE)
             )
p
```

## Multiple types in one table (2)

<div class='right' style='float:left;width:45%'>

Song Table

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
library(DT)
p<-datatable(song,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```

<div class='right' style='float:left;width:10%'>
```{r, echo=FALSE, out.width="90%", fig.cap="", fig.align = 'center'}
url = "https://www.svgrepo.com/show/3193/arrows-couple.svg"
knitr::include_graphics(url)
```
</div>

<div class='right' style='float:left;width:45%'>

Ranking Table

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../data/tidy-data/data/billboard.r")
library(DT)
p<-datatable(rank,rownames = FALSE,
             options = list(pageLength = 10,
            columnDefs = list(list(className = 'dt-left'))
            ,autoWidth = TRUE)
             )
p
```
</div>

## Examples

**Energy Audit**

This project was conducted in 2018 when I was with GHD. I was the PM for a consultancy project of about 1 million USD for 6 months. The project includes following scope
- Reliability check of all major equipment and systems of 20+ large pump stations in Metro Manila
- Energy Audit
- Conceptual design to upgrade the system

[https://namkyodai.github.io/post/2019-09-15-excel-combine/](https://namkyodai.github.io/post/2019-09-15-excel-combine/)

>>>>

---
title     : "Các Nguyên Lý Cơ Bản Trong Phân Tích Dữ Liệu"
subtitle  : "Làm Sạch và Xác Thực Dữ Liệu, Dữ Liệu Gọn Gàng"
author    : "[TS. Lê Thanh Nam](https://namkyodai.github.io)"
date      : "4-tháng 8-2023"
#date      : "`r format(Sys.time(), '%d %B, %Y')`"
output    : 
  slidy_presentation:
#   css: assets/css/styles.css
#   css: assets/css/ribbons.css
    mathjax: local
    self_contained: false
    font_adjustment: 0
    fig_width: 7
    fig_height: 6
    fig_caption: true
    incremental: false #for bullet increment
    highlight: pygments
    footer: "Bản Quyền của Cty TNHH ASQ Việt Nam - Chuẩn Bị cho Khóa Học tại Đại Học UEH"
    includes:
      in_header: assets/layouts/header.html
      #before_body: doc_prefix.html
      #after_body: doc_suffix.html
    duration: 45
---

## Làm Sạch và Xác Thực Dữ Liệu (1)

> Làm sạch dữ liệu là một quá trình phát hiện, sửa, loại bỏ, thay đổi các dòng dữ liệu không chính xác/ bị lỗi từ các bảng ghi, các cơ sở dữ liệu.  Công việc làm sạch có thể được tiến hành bởi các công cụ chuyên dụng, hoặc có thể được làm theo lô thông qua dòng lệnh.

>
> `r tufte::quote_footer('--- https://en.wikipedia.org/wiki/Data_cleansing')`

<span style="color: red;"> Các nhà khoa học dữ liệu sử dụng **60% thời gian của họ để làm sạch ** và thiết lập dữ liệu. Thu thập dữ liệu chiếm khoảng 19% thời gian, điều này có nghĩa là khoảng **80% thời gian của họ là dành cho việc chuẩn bị và quản lý dữ liệu**.</span>

> Xác thực dữ liệu là một hoạt động nhằm xác thực một tập hợp các giá trị là một thành viên của các tập hợp được chấp nhận.
>
> `r tufte::quote_footer('--- [Zio et al. 2016](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)')`

## Làm Sạch và Xác Thực Dữ Liệu (2)

@@

Source: [Phương Pháp Xác Thực Dữ Liệu 1.0](https://ec.europa.eu/eurostat/cros/system/files/methodology_for_data_validation_v1.0_rev-2016-06_final.pdf)

@@

- Xác Thực Cấp Độ 0: Tính nhất quán với các yêu cầu về cấu trúc IT
- Xác Thực Cấp Độ 1: Tính nhất quán trong chính gói dữ liệu
- Xác Thực Cấp Độ 2: Tính nhất quán với các gói dữ liệu trong cùng một miền và trong cùng một nguồn dữ liệu
- Xác Thực Cấp Độ 3:  Tính nhất quán trong cùng một miền với các nguồn dữ liệu khác nhau
- Xác Thực Cấp Độ 4:  Tính nhất quán giữa các miền tách biệt trong cùng một nhà cung cấp
- Xác Thực Cấp Độ 5: Tính nhất quán với dữ liệu của các nhà cung cấp.

@@

## Dữ Liệu Gọn Gàng - Cấu Trúc

@@

**Không Gọn Gàng**

Thể hiện điển hình

@@

Thể Hiện Điển Hình (Tiếp)

@@

- Cái gì là <span style="color: red;">**Biến**</span>, <span style="color: red;">**Quan sát**</span>, <span style="color: red;">**Giá trị**</span>?

@@

<div class='right' style='float:right;width:48%'>
> Các gia đình hạnh phúc thì đều giống nhau, các gia đình không hạnh phúc thì không hạnh phúc theo từng cách riêng của họ

>
> `r tufte::quote_footer('--- Leo Tolstoy')`

Như là các gia đình, các dữ liệu gọn gàng thì đều giống nhau, nhưng các dữ liệu lộn xộn, không gọn gàng thì lộn xộn theo từng cách riêng của nó và không giống nhau. Các gói dữ liệu gọn gàng cung cấp một cách tiêu chuẩn để kết nối cấu trúc của một tập dữ liệu (theo sơ đồ vật lý) và với nghữ nghĩa của nó (nghĩa).

@@

## Dữ Liệu Gọn Gàng - Định Nghĩa

@@

Một gói dữ liệu là một tập hợp các giá trị, thường ở dạng số hay dạng chữ. Các giá trị này được thiết lập theo 2 cách. Toàn bộ các giá trị thuộc về một **Biến** và một **Quan Sát**. Một biến chứa đựng trong nó toàn bộ các giá trị để đo lường cho cùng một thuộc tính (như chiều cao, nhiệt độ, thời gian) trên các đơn vị. Một Quan Sát chứa toàn bộ các giá trị được đo lường trên cùng một đơn vị (như một người, một ngày, một chủng tộng) trên các thuộc tính.

@@

Nguồn: [Openscapes](https://www.openscapes.org/blog/2020/10/12/tidy-data/)
</div>

@@

** Dữ liệu gọn gàng là một cách tiêu chuẩn của công việc chuyển thể (bản đồ hóa) ý nghĩa của một gói dữ liệu tới cấu trúc của nó. Một gói dữ liệu là lộn xộn hay gọn gàng phụ thuộc vào việc các dòng, các cột và các bảng của chúng có tương tích với các quan sát, các biến, và loại dữ liệu như thế nào**. Trong một dữ liệu gọn gàng thì:

  - Mỗi một biến cấu thành một cột.
  - Mỗi một quan sát cấu thành một hàng.
  - Mỗi một loại đơn vị quan sát cấu thành một bảng.

@@

## Dữ Liệu Gọn Gàng - Dọn Dẹp Phòng Của Bạn theo Cách của Chuyên Gia Marie Kondo (KONMARI)

@@

Gấp Đồ Lót Trong Tủ Quần Áo

@@

Thiết Kế Cuộc Sống Để Luôn Vui Vẻ

@@

## Dọn Dẹp Dữ Liệu Lộn Xộn (1)

- Tên các cột là giá trị, không phải là Biến
- Nhiều biến được ghi trong một cột

- Các biến được lưu ở cả hàng và cột

- Các loại dữ liệu của nhiều đối tượng quan sát được ghi trong cùng một bảng

- Một đối tượng quan sát được ghi ở nhiều bảng

@@

```{r echo=FALSE, message=FALSE,warning=FALSE}
source("../Code/pew-tidying.R")
#library(dplyr)
#library(reshape2)
#df=read.csv("../Data/tidy-data/data/pew-raw.csv", sep=",", header=TRUE)
</div>
